---
title: "occupancy-prototyping"
author: "JT Miller"
date: "2024-05-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Prototyping Occupancy Modeling on Desert Plants
The script: (data_process.R)[https://github.com/lmguzman/occ_historical/blob/main/analysis/odonates/scripts/data_process.R] by Vaughn was referenced to build this <br>

For this particular prototype, we're going to start out small with a perennial that is iconic in the Sonoran Desert: *Olneya tesota* also known as desert ironwood. <br>

Load Dependencies 
```{r}
library(tidyverse)
library(data.table)
library(sf)
library(nimble) # possibly remove
library(raster)
library(lubridate)
#library(reclin) # removed from CRAN
library(sp)
#library(rgdal) # removed from CRAN
library(geosphere) # possibly remove
```

#### General Outline for Occupancy Modeling as given in Shirey et al. 2023
1. Determine the proportion of "Community Sampling Events" (CSEs) for a given species *Olneya tesota*, where CSEs are defined as sampling events where more than 1 species is collected on the same day within a 1km of eachother. 50% is the threshold to arrive at the golden standard baseline for occupancy modeling approaches. 
2. Decide on the Spatial and Temporal Scale of the analysis, as well as the size of the occupancy intervals & visit intervals
3. Check the Range Overlap of the species 
4. Check the taxonomic scale and sampling methods. Should non-detections be inferred across the family and genus level? 
5. Check whether visit history increases, decreases, or remains constant through occupancy intervals. 
6. Remove sites that are only present for a _single_ occupancy interval. These will skew the visit history. 
7. If visit history decreases through time and the probability of CSEs is to low, re-evaluate from step 2 onwards. 

#### Set-up Data
Load Data
```{r}
# Downloaded GBIF and iDigBio occurrences for all NA plant species, then subsetted the data to just those within the Sonoran Desert EPA lvl3 ecoregion
sonoran_occ <- fread("/home/jtmiller/mybook_jt/desert-distribution-modeling/data/raw/sonoran-extent-plants.csv")

# Shapefiles containing the EPA lvl 3 ecoregions 
na_ecoregions_3 <- sf::read_sf("/home/jtmiller/mybook_jt/desert-distribution-modeling/data/raw/shapefiles/NA-level-3-epa-ecoregions/NA_CEC_Eco_Level3.shp")
```

Transform to coordinate proj to equal area, subset to just the Sonoran Desert for the question at hand. 
```{r}
crs_1 <- "+proj=aea +lat_1=20 +lat_2=60 +lat_0=40 +lon_0=-96 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m"
na_ecoregions_3 <- st_transform(na_ecoregions_3, crs = crs_1)
sonoran_shp <- subset(na_ecoregions_3, na_ecoregions_3$NA_L3NAME == "Sonoran Desert")
```

Apply a buffer to the edge of the shapefile. 
```{r}
sonoran_shp_b <- st_buffer(sonoran_shp, dist = 50000) # 50,000m = 50km 
```

Fix Some fields from the occurrence data, remove data missing geo-coords/those data with unreasonably high uncertainty.
```{r}
# fix coordinate/uncertainty fields class type
sonoran_occ <- sonoran_occ[, latitude := as.numeric(latitude)]
sonoran_occ <- sonoran_occ[, longitude := as.numeric(longitude)]
sonoran_occ <- sonoran_occ[, coordinateUncertaintyInMeters := as.numeric(coordinateUncertaintyInMeters)]
# remove records with NA coords 
sonoran_occ <- sonoran_occ[!is.na(latitude)]
sonoran_occ <- sonoran_occ[!is.na(longitude)]
# remove records with high coordinate uncertainty
sonoran_occ <- sonoran_occ[coordinateUncertaintyInMeters <= 50000] # must be less than or equal to the area of our grid cells used (50km)
```
Convert the occurrence data into spatial data (sf), then transform the coordinate projection to match the equal area of the shapefile projection
```{r}
occ_spat <- st_as_sf(sonoran_occ, coords = c("longitude", "latitude"), crs = "+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs", remove = FALSE)
occ_spat <- st_transform(occ_spat, crs = st_crs(sonoran_shp_b))
```

Subset the data to just the shapefile of the Sonoran shape (buffered)
```{r}
occ_spat <- occ_spat[sonoran_shp_b, ]
```

Give the data unique IDs per occurrence, then create a copy of the data with WGS84 proj for the CSEs calculations. 


